---
title: "Heart Attack Prediction"
author: "Trevor Okinda"
date: "2024"
output:
  github_document: 
    toc: yes
    toc_depth: 4
    fig_width: 6
    fig_height: 4
    df_print: default
editor_options:
  chunk_output_type: console
---

# Student Details

|                                              |     |
|----------------------------------------------|-----|
| **Student ID Number**                        | 134780 |
| **Student Name**                             | Trevor Okinda |
| **BBIT 4.2 Group**                           | C |
| **Project Name**                             | Heart Attack Prediction |

# Setup Chunk

**Note:** the following KnitR options have been set as the global defaults: <BR> `knitr::opts_chunk$set(echo = TRUE, warning = FALSE, eval = TRUE, collapse = FALSE, tidy = TRUE)`.

More KnitR options are documented here <https://bookdown.org/yihui/rmarkdown-cookbook/chunk-options.html> and here <https://yihui.org/knitr/options/>.

```{r setup, include=FALSE}
library(formatR)
knitr::opts_chunk$set(
  warning = FALSE,
  collapse = FALSE
)
```

### Source: 

The dataset that was used can be downloaded here: *\<https://www.kaggle.com/datasets/rashikrahmanpritom/heart-attack-analysis-prediction-dataset?select=heart.csv\>*

### Reference:

*\<Avagyan, Z. (2017). Weather CSV [Data set]. Kaggle. https://www.kaggle.com/datasets/zaraavagyan/weathercsv\>\
Refer to the APA 7th edition manual for rules on how to cite datasets: <https://apastyle.apa.org/style-grammar-guidelines/references/examples/data-set-references>*

# Exploratory Data Analysis
## Load dataset
```{r load dataset}
# Load necessary libraries
library(tidyverse)

# Load dataset
heart_attack_data <- read.csv("heart.csv", colClasses = c(
  age = "numeric",           # Age of the patient
  sex = "factor",            # Sex of the patient (0 = female, 1 = male)
  cp = "factor",             # Chest Pain type (1-4)
  trtbps = "numeric",        # Resting blood pressure (in mm Hg)
  chol = "numeric",          # Cholesterol in mg/dl
  fbs = "factor",            # Fasting blood sugar > 120 mg/dl (1 = true, 0 = false)
  restecg = "factor",        # Resting electrocardiographic results (0-2)
  thalachh = "numeric",      # Maximum heart rate achieved
  exng = "factor",           # Exercise induced angina (1 = yes, 0 = no)
  oldpeak = "numeric",       # Oldpeak
  slp = "factor",            # Slope of the peak exercise ST segment (0-2)
  caa = "factor",            # Number of major vessels (0-3)
  thall = "factor",          # Thall (1-3)
  o2_saturation = "numeric", # Oxygen saturation
  output = "factor"          # Target (0 = less chance of heart attack, 1 = more chance of heart attack)
))

# Convert categorical variables to factors with appropriate labels
heart_attack_data <- heart_attack_data %>%
  mutate(
    sex = factor(sex, levels = c(0, 1), labels = c("female", "male")),
    cp = factor(cp, levels = c(1, 2, 3, 4), labels = c("typical angina", "atypical angina", "non-anginal pain", "asymptomatic")),
    fbs = factor(fbs, levels = c(0, 1), labels = c("false", "true")),
    restecg = factor(restecg, levels = c(0, 1, 2), labels = c("normal", "ST-T wave abnormality", "left ventricular hypertrophy")),
    exng = factor(exng, levels = c(0, 1), labels = c("no", "yes")),
    slp = factor(slp, levels = c(0, 1, 2), labels = c("downsloping", "flat", "upsloping")),
    caa = factor(caa, levels = c(0, 1, 2, 3), labels = c("0", "1", "2", "3")),
    thall = factor(thall, levels = c(1, 2, 3), labels = c("normal", "fixed defect", "reversible defect")),
    output = factor(output, levels = c(0, 1), labels = c("less chance of heart attack", "more chance of heart attack"))
  )

# Display the structure of the dataset
str(heart_attack_data)

# View the first few rows of the dataset
head(heart_attack_data)

# View the dataset in a separate viewer window
View(heart_attack_data)
```

## Measures of Frequency
```{r MOF}
# Function to calculate and print frequency and relative frequency
calculate_frequencies <- function(data, column) {
  freq <- table(data[[column]])
  rel_freq <- prop.table(freq)
  
  freq_df <- data.frame(
    Category = names(freq),
    Frequency = as.numeric(freq),
    Relative_Frequency = as.numeric(rel_freq)
  )
  
  print(freq_df)
}

# Calculate and print frequencies for each categorical variable
categorical_vars <- c("sex", "cp", "fbs", "restecg", "exng", "slp", "caa", "thall", "output")

for (var in categorical_vars) {
  cat("\nFrequencies for", var, ":\n")
  calculate_frequencies(heart_attack_data, var)
}

```

## Measures of Central Tendency
```{r MOCT}
# Measures of Central Tendency
central_tendency <- heart_attack_data %>%
  summarize(
    mean_age = mean(age, na.rm = TRUE),
    median_age = median(age, na.rm = TRUE),
    mode_age = as.numeric(names(sort(table(age), decreasing = TRUE)[1])),
    
    mean_trtbps = mean(trtbps, na.rm = TRUE),
    median_trtbps = median(trtbps, na.rm = TRUE),
    mode_trtbps = as.numeric(names(sort(table(trtbps), decreasing = TRUE)[1])),
    
    mean_chol = mean(chol, na.rm = TRUE),
    median_chol = median(chol, na.rm = TRUE),
    mode_chol = as.numeric(names(sort(table(chol), decreasing = TRUE)[1])),
    
    mean_thalachh = mean(thalachh, na.rm = TRUE),
    median_thalachh = median(thalachh, na.rm = TRUE),
    mode_thalachh = as.numeric(names(sort(table(thalachh), decreasing = TRUE)[1])),
    
    mean_oldpeak = mean(oldpeak, na.rm = TRUE),
    median_oldpeak = median(oldpeak, na.rm = TRUE),
    mode_oldpeak = as.numeric(names(sort(table(oldpeak), decreasing = TRUE)[1])),
    
    mean_o2_saturation = mean(o2_saturation, na.rm = TRUE),
    median_o2_saturation = median(o2_saturation, na.rm = TRUE),
    mode_o2_saturation = as.numeric(names(sort(table(o2_saturation), decreasing = TRUE)[1]))
  )

print(central_tendency)
```

## Measures of Distribution
```{r MOD}
# Measures of Distribution
distribution_measures <- heart_attack_data %>%
  summarize(
    var_age = var(age, na.rm = TRUE),
    sd_age = sd(age, na.rm = TRUE),
    range_age = range(age, na.rm = TRUE),
    
    var_trtbps = var(trtbps, na.rm = TRUE),
    sd_trtbps = sd(trtbps, na.rm = TRUE),
    range_trtbps = range(trtbps, na.rm = TRUE),
    
    var_chol = var(chol, na.rm = TRUE),
    sd_chol = sd(chol, na.rm = TRUE),
    range_chol = range(chol, na.rm = TRUE),
    
    var_thalachh = var(thalachh, na.rm = TRUE),
    sd_thalachh = sd(thalachh, na.rm = TRUE),
    range_thalachh = range(thalachh, na.rm = TRUE),
    
    var_oldpeak = var(oldpeak, na.rm = TRUE),
    sd_oldpeak = sd(oldpeak, na.rm = TRUE),
    range_oldpeak = range(oldpeak, na.rm = TRUE),
    
    var_o2_saturation = var(o2_saturation, na.rm = TRUE),
    sd_o2_saturation = sd(o2_saturation, na.rm = TRUE),
    range_o2_saturation = range(o2_saturation, na.rm = TRUE)
  )

print(distribution_measures)
```

## Measures of Relationship
```{r MOR}
# Measures of Relationship
relationship_measures <- heart_attack_data %>%
  summarize(
    cor_age_trtbps = cor(age, trtbps, use = "complete.obs"),
    cor_age_chol = cor(age, chol, use = "complete.obs"),
    cor_age_thalachh = cor(age, thalachh, use = "complete.obs"),
    cor_age_oldpeak = cor(age, oldpeak, use = "complete.obs"),
    cor_age_o2_saturation = cor(age, o2_saturation, use = "complete.obs"),
    
    cor_trtbps_chol = cor(trtbps, chol, use = "complete.obs"),
    cor_trtbps_thalachh = cor(trtbps, thalachh, use = "complete.obs"),
    cor_trtbps_oldpeak = cor(trtbps, oldpeak, use = "complete.obs"),
    cor_trtbps_o2_saturation = cor(trtbps, o2_saturation, use = "complete.obs"),
    
    cor_chol_thalachh = cor(chol, thalachh, use = "complete.obs"),
    cor_chol_oldpeak = cor(chol, oldpeak, use = "complete.obs"),
    cor_chol_o2_saturation = cor(chol, o2_saturation, use = "complete.obs"),
    
    cor_thalachh_oldpeak = cor(thalachh, oldpeak, use = "complete.obs"),
    cor_thalachh_o2_saturation = cor(thalachh, o2_saturation, use = "complete.obs"),
    
    cor_oldpeak_o2_saturation = cor(oldpeak, o2_saturation, use = "complete.obs")
  )

print(relationship_measures)
```

## ANOVA
```{r ANOVA}
# ANOVA for age across different levels of chest pain type (cp)
anova_age_cp <- aov(age ~ cp, data = heart_attack_data)
summary(anova_age_cp)

# ANOVA for cholesterol (chol) across different levels of sex
anova_chol_sex <- aov(chol ~ sex, data = heart_attack_data)
summary(anova_chol_sex)

# ANOVA for maximum heart rate achieved (thalachh) across different levels of output
anova_thalachh_output <- aov(thalachh ~ output, data = heart_attack_data)
summary(anova_thalachh_output)
```

## Plots
```{r Plots}
# Histogram for Age
ggplot(heart_attack_data, aes(x = age)) +
  geom_histogram(binwidth = 5, fill = "blue", color = "black", alpha = 0.7) +
  labs(title = "Histogram of Age", x = "Age", y = "Frequency")

# Density Plot for Cholesterol
ggplot(heart_attack_data, aes(x = chol)) +
  geom_density(fill = "green", alpha = 0.7) +
  labs(title = "Density Plot of Cholesterol", x = "Cholesterol (mg/dl)", y = "Density")

# Boxplot for Resting Blood Pressure
ggplot(heart_attack_data, aes(y = trtbps)) +
  geom_boxplot(fill = "orange", color = "black", alpha = 0.7) +
  labs(title = "Boxplot of Resting Blood Pressure", y = "Resting Blood Pressure (mm Hg)")

# Scatter Plot of Age vs. Maximum Heart Rate, colored by Chest Pain Type
ggplot(heart_attack_data, aes(x = age, y = thalachh, color = cp)) +
  geom_point(alpha = 0.7) +
  labs(title = "Scatter Plot of Age vs. Maximum Heart Rate", x = "Age", y = "Maximum Heart Rate (thalachh)") +
  scale_color_discrete(name = "Chest Pain Type")

# Boxplot of Cholesterol by Sex
ggplot(heart_attack_data, aes(x = sex, y = chol, fill = sex)) +
  geom_boxplot(alpha = 0.7) +
  labs(title = "Boxplot of Cholesterol by Sex", x = "Sex", y = "Cholesterol (mg/dl)") +
  scale_fill_discrete(name = "Sex")

# Pair Plot of Selected Numeric Variables
# Load GGally library for ggpairs function
library(GGally)

# Select numeric columns for pair plot
numeric_columns <- heart_attack_data %>%
  select(age, trtbps, chol, thalachh, oldpeak, o2_saturation)

# Create pair plot
ggpairs(numeric_columns, 
        title = "Pair Plot of Selected Numeric Variables")

```

# Preprocessing and Data Transformation
## Missing Values
```{r Missing Values}
# 1. Check for missing values in each column
missing_values <- colSums(is.na(heart_attack_data))
print(missing_values)

# 2. Summary of missing values across the dataset
missing_summary <- summary(is.na(heart_attack_data))
print(missing_summary)

library(naniar)  # For visualizing missing data
# 3. Visual representation of missing values
gg_miss_var(heart_attack_data) + 
  labs(title = "Missing Values in Heart Attack Dataset", x = "Variables", y = "Number of Missing Values")

# Another visual representation using visdat package
# Load visdat package
library(visdat)

# Visualize missing values
vis_miss(heart_attack_data) + 
  labs(title = "Missing Values in Heart Attack Dataset")

#Remove rows with missing values
heart_attack_data_clean <- na.omit(heart_attack_data)

#Summary of missing values across the cleaned dataset
missing_summary <- summary(is.na(heart_attack_data_clean))
print(missing_summary)
```

# Training Model
## Data Splitting
```{r Data Splitting}
# Load necessary library
library(caret)

# Set seed for reproducibility
set.seed(123)

# Split the data into training (70%) and testing (30%) sets
train_index <- createDataPartition(heart_attack_data_clean$output, p = 0.7, list = FALSE)
train_data <- heart_attack_data_clean[train_index, ]
test_data <- heart_attack_data_clean[-train_index, ]

# Print the dimensions of the training and testing sets
cat("Training set dimensions:", dim(train_data), "\n")
cat("Testing set dimensions:", dim(test_data), "\n")
```

## Cross-validation
```{r Cross-validation}
# Set seed for reproducibility
set.seed(123)

# Define cross-validation control
cv <- trainControl(method = "cv", number = 5, verboseIter = TRUE)

# Your modeling code here (e.g., train a model using caret's train function)
# Example: train a logistic regression model using 5-fold cross-validation
model <- train(output ~ ., data = train_data, method = "glm", family = "binomial", trControl = cv)

# Print model performance
print(model)
```

## Train different models
```{r Train different models}
# Load necessary libraries
library(randomForest)
library(e1071)  # For SVM
library(caret)

# Define training control
ctrl <- trainControl(method = "cv", number = 5, verboseIter = TRUE)

# Train Random Forest model
set.seed(123)
rf_model <- train(output ~ ., data = heart_attack_data_clean, method = "rf", trControl = ctrl)

# Train Support Vector Machine (SVM) model
set.seed(123)
svm_model <- train(output ~ ., data = heart_attack_data_clean, method = "svmRadial", trControl = ctrl)

# Train Logistic Regression model
set.seed(123)
logit_model <- train(output ~ ., data = heart_attack_data_clean, method = "glm", family = "binomial", trControl = ctrl)

# Print model results
print(rf_model)
print(svm_model)
print(logit_model)
```

## Performance Comparison
```{r Performance Comparison}
# Load necessary libraries
library(caret)

# Define resampling method
ctrl <- trainControl(method = "cv", number = 5, verboseIter = TRUE)

# Compare model performance
models <- list(rf = rf_model, svm = svm_model, logit = logit_model)
model_resamples <- resamples(models)

# Summarize results
summary(model_resamples)
```

## Bootstrapping
```{r Bootstrapping}
# Set seed for reproducibility
set.seed(123)

# Number of bootstrap iterations
n_bootstraps <- 1000

# Bootstrap function for age variable
bootstrap_age <- replicate(n_bootstraps, {
  # Sample with replacement from the age variable
  bootstrap_sample <- sample(heart_attack_data_clean$age, replace = TRUE)
  
  # Calculate statistic of interest (e.g., mean, median, standard deviation) for each bootstrap sample
  # Mean and standard deviation
  mean_age <- mean(bootstrap_sample)
  sd_age <- sd(bootstrap_sample)
  
  return(c(mean_age, sd_age))
})

# Calculate mean and standard deviation for each bootstrap sample
bootstrap_stats <- t(bootstrap_age)

# Calculate confidence intervals for mean and standard deviation
confidence_intervals <- apply(bootstrap_stats, 2, function(stat) {
  quantile(stat, c(0.025, 0.975))
})

# Print results
cat("Bootstrap Mean Age:", mean(bootstrap_stats[, 1]), "\n")
cat("95% Confidence Interval for Mean Age:", confidence_intervals[1, 1], "-", confidence_intervals[2, 1], "\n")
cat("Bootstrap Standard Deviation Age:", mean(bootstrap_stats[, 2]), "\n")
cat("95% Confidence Interval for Standard Deviation Age:", confidence_intervals[1, 2], "-", confidence_intervals[2, 2], "\n")

```

## Saving Model
```{r Saving Model}

# Saving the Random Forest model for the heart attack dataset
saveRDS(rf_model, file = "./models/heart_attack_rf_model.rds")

# Load the saved Random Forest model
loaded_heart_attack_rf_model <- readRDS("./models/heart_attack_rf_model.rds")

# Prepare new data for prediction (adjusting for factor conversions)
new_heart_attack_data <- data.frame(
  age = c(63, 37, 41),   # Example age values
  sex = factor(c(1, 1, 0), levels = c(0, 1), labels = c("female", "male")),  # Adjusted sex values
  cp = factor(c(3, 2, 1), levels = c(1, 2, 3, 4), labels = c("typical angina", "atypical angina", "non-anginal pain", "asymptomatic")),  # Adjusted cp values
  trtbps = c(145, 130, 130),  # Example trtbps values
  chol = c(233, 250, 204),    # Example chol values
  fbs = factor(c(1, 0, 0), levels = c(0, 1), labels = c("false", "true")),      # Adjusted fbs values
  restecg = factor(c(0, 1, 0), levels = c(0, 1, 2), labels = c("normal", "ST-T wave abnormality", "left ventricular hypertrophy")),  # Adjusted restecg values
  thalachh = c(150, 187, 172),  # Example thalachh values
  exng = factor(c(0, 0, 0), levels = c(0, 1), labels = c("no", "yes")),     # Adjusted exng values
  oldpeak = c(2.3, 3.5, 1.4),   # Example oldpeak values
  slp = factor(c(0, 0, 2), levels = c(0, 1, 2), labels = c("downsloping", "flat", "upsloping")),      # Adjusted slp values
  caa = factor(c(0, 0, 2), levels = c(0, 1, 2, 3), labels = c("0", "1", "2", "3")),      # Adjusted caa values
  thall = factor(c(1, 2, 2), levels = c(1, 2, 3), labels = c("normal", "fixed defect", "reversible defect")),    # Adjusted thall values
  o2_saturation = c(98.6, 98.6, 98.6)  # Example o2_saturation values
)

# Use the loaded model to make predictions for new heart attack data
predictions_rf_loaded_model <- predict(loaded_heart_attack_rf_model, newdata = new_heart_attack_data)

# Print predictions
print(predictions_rf_loaded_model)


```

